---
title: "Readme"
author: "Hanna McCaslin"
date: "6/2/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This repository contains the code and data for converting individual-by-individual BBL records to MARK frequency data. 

The data from the BBL in May are attached in the 'data' folder in the repository (hopefully). There's also a location file to match up the individual records to a place, just in case. 
```{r, include = FALSE}
source("R/bbl-functions.R")

### Read in encounter data and banding data from BBL (downloaded from ftp May 2017)
encounters <- read_bbl("data/AMKE_encounterdb_201705111648.csv") %>%
     select(BAND_NUM, B_SEX_CODE, B_AGE_CODE, B_FLYWAY_CODE, BANDING_YEAR, E_HOW_OBTAINED_CODE, E_PRESENT_CONDITION_CODE, ENCOUNTER_YEAR, B_REGION, B_LON_DECIMAL_DEGREES)

all_bands <- read_bbl("data/AMKE_bandingdb_201705111647.csv", with_encs = FALSE) %>%
     select(BAND_NUM, B_SEX_CODE = SEX_CODE, B_AGE_CODE = AGE_CODE, B_FLYWAY_CODE = FLYWAY_CODE, BANDING_YEAR, B_REGION = REGION, B_LON_DECIMAL_DEGREES = LON_DECIMAL_DEGREES)
```

## Filtering encounter records
Originally, I just filtered down to only include present condition codes 03, 04, 05 which are the "dead" codes:
```{r}
recoveries <- filter(encounters, E_PRESENT_CONDITION_CODE == 3 | E_PRESENT_CONDITION_CODE == 4 | E_PRESENT_CONDITION_CODE == 5)
# results in encounters n = 3484
```

I then went back and refiltered using the present condition/how obtained code combinations Julie used, which results in the inclusion of 73 additional records:
```{r}
recoveries2 <- filter(encounters, E_PRESENT_CONDITION_CODE != 06 & E_PRESENT_CONDITION_CODE != 07 & E_PRESENT_CONDITION_CODE != 08) #removes 1508
recoveries2 <- filter(recoveries2, E_PRESENT_CONDITION_CODE == 03 | E_PRESENT_CONDITION_CODE == 04 | E_PRESENT_CONDITION_CODE == 05 | (E_HOW_OBTAINED_CODE != 03 & E_HOW_OBTAINED_CODE != 21 & E_HOW_OBTAINED_CODE != 28 & E_HOW_OBTAINED_CODE != 29 & E_HOW_OBTAINED_CODE != 33 & E_HOW_OBTAINED_CODE != 36 & E_HOW_OBTAINED_CODE != 37 & E_HOW_OBTAINED_CODE != 46 & E_HOW_OBTAINED_CODE != 47 & E_HOW_OBTAINED_CODE != 53 & E_HOW_OBTAINED_CODE != 56 & E_HOW_OBTAINED_CODE != 66 & E_HOW_OBTAINED_CODE != 97 & E_HOW_OBTAINED_CODE != 98 ))

filter(recoveries2, BAND_NUM == 126346015) %>% # One duplicate - both encounters report the bird dead, but in different columns
     select(ENCOUNTER_YEAR, E_HOW_OBTAINED_CODE, E_PRESENT_CONDITION_CODE)
recoveries2 <- filter(recoveries2, BAND_NUM != 126346015 | E_PRESENT_CONDITION_CODE ==05) #remove one of the duplicates
# results in n = 3557
```

One thing that's important to note here is that I'm just removing these from the encounter set, which next I will join with the banding records by band number, so the individuals taken out aren't gone from the dataset; they are just treated as surviving birds. The problem with this is that some of the records that got removed/pushed into the surviving birds probably shouldn't be considered alive or dead at all and should be removed altogether from the data. Does that make sense?  
What do you think - which filtering do you want to use, and should I go back through and remove some of the more ambiguous individuals from the data altogether?  

## Joining encounters with banding records and filtering out records
After joining the two datasets together, any filtering will actually remove that record/band from the data altogether. (This distinction is something I missed initially, and could explain some of the difference in number of records)

```{r, include = FALSE}
bands_and_recoveries <- full_join(all_bands, recoveries, by = c("BAND_NUM"))

unfiltered_data <- select(bands_and_recoveries, BAND_NUM, SEX = B_SEX_CODE.x, AGE = B_AGE_CODE.x, FLYWAY = B_FLYWAY_CODE.x, BANDING_YEAR = BANDING_YEAR.x, B_REGION = B_REGION.x, B_LON = B_LON_DECIMAL_DEGREES.x, ENCOUNTER_YEAR, E_HOW_OBTAINED = E_HOW_OBTAINED_CODE, E_CONDITION = E_PRESENT_CONDITION_CODE)
```

This is the filtering that I ended up doing: removing unknown age and sex, removing Caribbean and South American records, removing Mexico, selecting the correct years, and removing the locals that died in the nest. 
```{r}
filtered_data <- unfiltered_data %>%
     filter(SEX != 0) %>%  # remove unknown sex
     filter(AGE != 0) %>%  # remove unknown age
     filter(FLYWAY != 8 & FLYWAY != 9) %>% # remove Caribbean & South American records 
     filter(FLYWAY!= 7) %>% # optional: remove mexico
     filter(BANDING_YEAR >= 1967 & BANDING_YEAR <= 2015) %>% 
     filter(is.na(ENCOUNTER_YEAR) | (ENCOUNTER_YEAR >= 1967 & ENCOUNTER_YEAR <= 2015)) %>% # filter to dates of study
     filter(is.na(E_HOW_OBTAINED) | !(AGE == 4 & E_HOW_OBTAINED == 30)) #remove locals died in nest, removes n = 19

     #n = 267118 / 266990 remaining (Mex, no Mex)
```

And, I did the same with the encounters selected by Julie's specific criteria:
```{r, include = FALSE}
bands_and_recoveries2 <- full_join(all_bands, recoveries2, by = c("BAND_NUM")) 

unfiltered_data2 <- select(bands_and_recoveries2, BAND_NUM, SEX = B_SEX_CODE.x, AGE = B_AGE_CODE.x, FLYWAY = B_FLYWAY_CODE.x, BANDING_YEAR = BANDING_YEAR.x, B_REGION = B_REGION.x, B_LON = B_LON_DECIMAL_DEGREES.x, ENCOUNTER_YEAR, E_HOW_OBTAINED = E_HOW_OBTAINED_CODE, E_CONDITION = E_PRESENT_CONDITION_CODE)
```

```{r}
filtered_data2 <- unfiltered_data2 %>%
     filter(SEX != 0) %>%  # remove unknown sex
     filter(AGE != 0) %>%  # remove unknown age
     filter(FLYWAY != 8 & FLYWAY != 9) %>% # remove Caribbean & South American records
     filter(FLYWAY!= 7) %>% #optional: remove mexico
     filter(BANDING_YEAR >= 1967 & BANDING_YEAR <= 2015) %>% 
     filter(is.na(ENCOUNTER_YEAR) | (ENCOUNTER_YEAR >= 1967 & ENCOUNTER_YEAR <= 2015)) %>%  # filter to dates of study
     filter(is.na(E_HOW_OBTAINED) | !(AGE == 04 & E_HOW_OBTAINED == 30))

#n = 267115 / 266987 remaining (Mex, no Mex)
```

So the filtering doesn't turn out to matter too much, because there are only three records that stay in filtering one way or the other:
```{r, include = FALSE}
filter(filtered_data, BAND_NUM == 178385863 | BAND_NUM == 162335894 | BAND_NUM == 178317149)
     
```

### Without HY
Using the originally filtered data (I'd be happy to take out those other 3 records), here's the data with out the hatch years. I actually don't think you eliminated all of these in the first go because it makes us about 10K records short of where you were.
```{r}
data_noHY <- filter(filtered_data, AGE != 02) # removes n = 73871
```

```{r, include = FALSE}
### Transform some columns ----
# Sex (into M and F)
data_chr_traits <- translate_sex(filtered_data, "SEX")
data_chr_noHY <- translate_sex(data_noHY, "SEX")

# Age (into HY and AHY categories)
data_chr_traits <- translate_age(data_chr_traits, "AGE", group = TRUE) 
data_chr_noHY <- translate_age(data_chr_noHY, "AGE", group = TRUE)

#Flyway (into 4)
data_chr_traits <- calc_flyways_four(data_chr_traits, "FLYWAY", "B_LON", chr_codes = FALSE, int_codes = TRUE) 
data_chr_noHY <- calc_flyways_four(data_chr_noHY, "FLYWAY", "B_LON", chr_codes = FALSE, int_codes = TRUE)

#Get rid of some unneccessary columns
amke_individuals <- select(data_chr_traits, BAND_NUM, sex, age_class = b_age, flyway = FLYWAY, b_year = BANDING_YEAR, e_year = ENCOUNTER_YEAR)
amke_indivs_noHY <- select(data_chr_noHY, BAND_NUM, sex, age_class = b_age, flyway = FLYWAY, b_year = BANDING_YEAR, e_year = ENCOUNTER_YEAR)
```

Another place that you could have lost some records was with the flyway filtering. When I used the longitudes exactly as you gave me, I lost 208 records from Canada that were farther east/west than the boundaries for flyways 1/4, so I changed the 4 flyways to span all of Canada. I also made the boundaries touch each other just because. I can change either of these things back too, just let me know.
    
     fw4 <- c(-141, -110) [what I used]          fw4 <- c(-136, -110) [what you used]
     fw3 <- c(-110, -90)
     fw2 <- c(-90, -82.7)
     fw1 <- c(-82.7, -52)                        fw1 <- c(-82.7, -59)

## Final data sets

```{r}
make_ldld <- function(df, study_first_yr, study_last_yr){
     
     reference_year1 <- study_first_yr
     reference_year2 <-study_last_yr
     
     study_length <- (reference_year2 - reference_year1) + 1
     
     #Create a string of 0's of length 2 * study length
     ch <- ""
     m = 0
     while(m < 2 * study_length){
          ch <- str_c(ch, "0")
          m <- m + 1
     }

     band_position <- vector(mode = "integer", length = nrow(df))
     enc_position <- vector(mode = "integer", length = nrow(df))
     result <- vector(mode = "character", length = nrow(df))
     
     for(i in 1:nrow(df)){result[i] = ch} #fill result with character strings
     for(i in 1:nrow(df)){
          band_position[i] <- 2*(df$b_year[i] - (reference_year1-1))-1 # create a vector of all the places the band "1" will go
     }
     for(i in 1:nrow(df)){
          str_sub(result[i], start = band_position[i], end = band_position[i]) <- "1" # update the strings in result with band "1"
     }
     for(i in 1:nrow(df)){
          if(!is.na(df$e_year[i])){
               enc_position[i] <- 2*(df$e_year[i] - (reference_year1-1)) # create a vector of all the places the enc "1" will go
          } else{
               enc_position[i] <- NA
          }
     }
     for(i in 1:nrow(df)){
          if(!is.na(enc_position[i])){
               str_sub(result[i], start = enc_position[i], end = enc_position[i]) <- "1"
          }
     }
     
     result <- str_c("\"", result, "\"") #add " to the ends of the string so excel doesn't eat the 0s
     df$ch <- result
     return(df)
}
```

I made versions including and excluding HYs
```{r}
### Make mark string in dataset
amke_mark <- make_ldld(amke_individuals, 1967, 2015)
amke_noHY_mark <- make_ldld(amke_indivs_noHY, 1967, 2015)

### Group into rows w/ frequencies
grouped_amke <- group_by(amke_mark, ch, sex, age_class, flyway)
final <- summarise(grouped_amke, freq = n())
View(final)
sum(final$freq)

#And again for no-HYs
grouped_amke_noHY <- group_by(amke_noHY_mark, ch, sex, age_class, flyway)
final_noHY <- summarise(grouped_amke_noHY, freq = n())
View(final_noHY)
sum(final_noHY$freq)
```

What's interesting here that I can't totally figure out is that I have about 300 more possible LD/age/sex combinations (rows) than Ben had when he did it, even when I take out the hatch years and have a smaller dataset. Ideas?

## Files
The csvs of the end data with and without hatch years are (hopefully) in the mark-data file in the repository. I'll try to be sure to update them if I change anything.  

As of 6 June 2017, the data files are:  
     "mark-data/LD_amke_may2017.csv" for with HY   
     "mark-data/LD_amke_noHY_may2017.csv" for without HY
